Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2016-12-01 draft.tex
.

\begin{document}
\maketitle
\section{Abstract} \label{abstract}
Based on the LIF-SORN-model proposed in \cite{SORN_Paper}, we attempted to replace the intrinsic homeostatic control system used in the original version by a mechanism based on the diffusion of a neurotransmitter across the nervous tissue. The model of diffusive homeostasis was adopted from a paper by Y. Sweeney et al. \cite{Sweeney_Paper} and models the tissue as a surface of square shape and a set of points on this surface, representing the neurons' positions within the SORN. The group of excitatory neurons then acted as a point-source of nitric oxide (NO), as well as as a sensor for the NO-concentration at each individual position. The production and sensing of NO forms the basis of a feedback loop: The individual NO-readout is fed into a comparator which causes an appropriate change within the internal firing threshold of the neuron, in turn altering the neuron's firing rate. The control system is then closed by linking the rate of NO-production to the neuron's firing rate.

@@

\subsection{Analysis of Oscillations under Dynamic Feedback}\label{theor_osc}
In this section, we discuss an analytic treatment of the dynamics underlying the diffusive homeostatic mechanism. The goal of this analysis is to predict the shape of the power spectrum shown in Figure \ref{Power_Spec_without_Analysis}. This will give insight into the relation between parameters of the model and the resulting preferred frequency and amplitude, allowing us to choose appropriate parameter values in order to reduce frequency and amplitude of the oscillations.

Both forms of homeostasis use the excitatory firing threshold as a means of adjusting the excitatory firing rate. Though having a relatively immediate impact on the firing rate of the particular neuron, the network as a whole reacts by means of a local disturbance of activity as well: it settles at a new fixed point of firing rates. In the context of feedback dynamics, two questions are of importance. First, obviously, how the relation between a local change of threshold and the new fixed point of the network can be described and second, whether the time scale of the network's response is of relevance with respect to other time scales involved in the feedback loop. Regarding the second question, Figure \ref{rate_th_close} suggests that the excitatory population activity follows the mean firing threshold quasi instantaneously, compared with the overall time scale of the oscillating pattern.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/diff_hom/rate_th_close.png}
\end{center}
\caption{Time course of the mean firing rate (blue) and the mean firing threshold within the excitatory pool (red). Bin width for firing rate estimation was 0.1s.}
\label{rate_th_close}
\end{figure}
Thus, in a first attempt of understanding the occurrence of undesired oscillations, we presumed an immediate functional relationship $\mathbf{r}_e (t) = \mathbf{r}_e (\mathbf{V}_{t,e} (t))$, $\mathbf{r}_e$ and $\mathbf{V}_{t,e}$ representing the set of excitatory firing rates and thresholds, respectively.

@@

To overcome this contradiction, we introduce an idealized picture of diffusive homeostasis, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses and the decay term $-\lambda NO$. To account for the spread of NO, $nNOS(t)$ must be divided by the area $L^2$ of the tissue. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\sum_{i} nNOS^i(t)}{L^2}\;.
\label{NO_dyn_approx}
\end{equation}
In this form, it becomes clear that we can now describe the \emph{total} influx of NO (the sum in \eqref{NO_dyn_approx}) as the sum of a mean and a random fluctuation: While spiking events on individual cells remain sufficiently separated to justify equation \eqref{nNOS_single_sol_approx2}, the sum of all spike events from the excitatory population results in an effective poisson process with a mean rate of $3 Hz \cdot N_{exc.} = \mathrm{1200\,Hz}$, given the assumption that cells are spiking weakly correlated. Furthermore, we argue that a total rate of $\mathrm{1200\,Hz}$ is sufficiently large to express this rate as a continuous function of time, $r_{total} = N_{exc.}\cdot r_{exc.pop.}(t)$, at least on the time scale of the oscillations to be studied (see  Figure \ref{rate_th_close}). Thus, equation \eqref{NO_dyn_approx} becomes
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\mu_{nNOS}(N_{exc.}r_{exc.pop.}(t)) + \sigma_{nNOS}(N_{exc.}r_{exc.pop.}(t)) \xi (t)}{L^2}\;.
\label{NO_dyn_approx2}
\end{equation}
As a next step, we attempted to simplify the dynamics of threshold adaption and its effect on the population rate. Since equation \eqref{NO_dyn_approx2} contains a single variable $r_{exc.pop.}(t)$ for the mean excitatory activity, we wanted to find an appropriate description containing the mean threshold $V_{t,pop.}(t) = <V_t(t)>$. As stated earlier, Figure \ref{rate_th_close} suggests an immediate and approximately linear relation between excitatory population rate and the mean of thresholds within the excitatory population. Figure \ref{thresh_r_linfit} shows a linear fit (being of the form $\alpha x + \beta$) of these two quantities plotted against each other. Apart from rarely appearing high values of $V_{t,pop.}$, we found a good fit to the linear model and concluded $r_{exc.pop.}(t) \approx r_{IP} + \alpha(V_{t,pop.}(t)-V_{t,0})$, where $r_{IP}$ is given in table \ref{Params_IP} as the mean excitatory target firing rate and $V_{t,0} \equiv (r_{IP}-\beta)/\alpha$ is the mean threshold corresponding to the target rate.
\begin{figure}
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/linfit_th_rate.png}
\caption{Linear fit of excitatory population activity and mean excitatory threshold. Slope $\alpha$ of the fit: $\mathrm{-0.772\, Hz/mV}$. Offset $\beta$: $\mathrm{-41.01\, Hz}$. Mean squared error: $\mathrm{R^2 = 0.69}$.}
\label{thresh_r_linfit}
\end{figure}

Combining these results in a set of equations, we find
\begin{align}
\begin{split}
\dot{NO}(t) &= - \lambda NO + \frac{ N_{exc.} {Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3 L^2}[\alpha(V_{t,pop.}(t)-V_{t,0}) + r_{IP}] \\
&+ \frac{{Ca^{2+}_{spike}}^3 {\tau_{Ca^{2+}}} ln(2)}{3L^2}\sqrt{\frac{N_{exc.}}{2 \tau_{nNOS}}}\sqrt{\alpha(V_{t,pop.}(t)-V_{t,0}) + r_{IP}}\xi (t)
\end{split}\label{2d-sys-homeostasis1}\\
\dot{V_{t,pop.}}(t) &= \frac{NO-NO_0}{NO_0 \tau_{V_t}}\;. \label{2d-sys-homeostasis2}
\end{align}
Through the coordinate transformation $n = NO-NO_0 \; , \theta = V_{t,pop.}(t)-V_{t,0}$, we can simplify the dynamical system:
\begin{align}
\begin{split}
\dot{n} &= -\lambda n + \frac{\alpha N_{exc.} {Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3 L^2}\theta \\
&+ \frac{{Ca^{2+}_{spike}}^3 {\tau_{Ca^{2+}}} ln(2)}{3L^2}\sqrt{\frac{N_{exc.}}{2 \tau_{nNOS}}}\sqrt{\alpha\theta + r_{IP}}\xi (t)
\end{split}\label{2d-sys-homeostasis-coord_transf1}\\
\dot{\theta} &= \frac{1}{NO_0 \tau_{V_t}} n \label{2d-sys-homeostasis-coord_transf2}
\end{align}
The amplitude of noise in equation \ref{2d-sys-homeostasis-coord_transf1} depends on $\theta$. If kept in the equation, this dependence turned out be an insuperable obstacle for our goal of finding an analytic description of the homeostatic oscillations. In consequence, we decided to make one further simplification, replacing $\sqrt{\alpha\theta + r_{IP}}$ by a constant factor $\sqrt{r_{IP}}$.

>>>>

\begin{document}
\maketitle
\section{Abstract} \label{abstract}
We attempted to replace the intrinsic homeostatic control system used in the original version by a mechanism based on the diffusion of a neurotransmitter across the nervous tissue based on the LIF-SORN-model proposed in \cite{SORN_Paper}. The model of diffusive homeostasis was adopted from a paper by Y. Sweeney et al. \cite{Sweeney_Paper} and models the tissue as a surface of square shape and a set of points on this surface, representing the neurons' positions within the SORN. The group of excitatory neurons then acted as a point-source of nitric oxide (NO), as well as as a sensor for the NO-concentration at each individual position. The production and sensing of NO forms the basis of a feedback loop: The individual NO-readout is fed into a comparator which causes an appropriate change within the internal firing threshold of the neuron, in turn altering the neuron's firing rate. The control system is then closed by linking the rate of NO-production to the neuron's firing rate.

@@

\subsection{Analysis of Oscillations under Dynamic Feedback}\label{theor_osc}
In this section, we discuss an analytic treatment of the dynamics underlying the diffusive homeostatic mechanism. The goal of this analysis was to predict the shape of the power spectrum shown in Figure \ref{Power_Spec_without_Analysis}. This gave insight into the relation between parameters of the model and the resulting preferred frequency and amplitude, allowing us to choose appropriate parameter values in order to reduce frequency and amplitude of the oscillations.

Both forms of homeostasis use excitatory firing thresholds as a means of adjusting the excitatory firing rate. Though having a relatively immediate impact on the firing rate of the particular neuron, the network as a whole reacts by means of a local disturbance of activity as well: it settles at a new fixed point of firing rates. Two questions are of importance in the context of feedback dynamics. First - obviously - how the relation between a local change of threshold and the new fixed point of the network can be described and second, whether the time scale of the network's response is of relevance with respect to other time scales involved in the feedback loop. Regarding the second question, Figure \ref{rate_th_close} suggests that the excitatory population activity follows the mean firing threshold in a quasi instantaneous fashion, at least in comparison with the overall time scale of the oscillating pattern.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/diff_hom/rate_th_close.png}
\end{center}
\caption{Time course of the mean firing rate (blue) and the mean firing threshold within the excitatory pool (red). Bin width for firing rate estimation was 0.1 s.}
\label{rate_th_close}
\end{figure}
Thus, in a first attempt of understanding the occurrence of undesired oscillations, we presumed an immediate functional relationship $\mathbf{r}_e (t) = \mathbf{r}_e (\mathbf{V}_{t,e} (t))$, $\mathbf{r}_e$ and $\mathbf{V}_{t,e}$ representing the set of excitatory firing rates and thresholds, respectively.

@@

To overcome this contradiction, we introduce an idealized picture of diffusive homeostasis, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses and the decay term $-\lambda NO$. To account for the spread of NO, $nNOS(t)$ must be divided by the area $L^2$ of the tissue. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\sum_{i} nNOS^i(t)}{L^2}\;.
\label{NO_dyn_approx}
\end{equation}
In this form, it becomes clear that we can now describe the \emph{total} influx of NO (the sum in \eqref{NO_dyn_approx}) as the sum of a mean and a random fluctuation: While spiking events on individual cells remain sufficiently separated to justify equation \eqref{nNOS_single_sol_approx2}, the sum of all spike events from the excitatory population results in an effective poisson process with a mean rate of $3 Hz \cdot N_{exc.} = \mathrm{1200\,Hz}$, given the assumption that cells are spiking weakly correlated. Furthermore, we argue that a total rate of $\mathrm{1200\,Hz}$ is sufficiently large to express this rate as a continuous function of time, $r_{total} = N_{exc.}\cdot r_{exc.pop.}(t)$, at least on the time scale of the oscillations to be studied (see  Figure \ref{rate_th_close}). Thus, equation \eqref{NO_dyn_approx} becomes
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\mu_{nNOS}(N_{exc.}r_{exc.pop.}(t)) + \sigma_{nNOS}(N_{exc.}r_{exc.pop.}(t)) \xi (t)}{L^2}\;.
\label{NO_dyn_approx2}
\end{equation}
As a next step, we attempted to simplify the dynamics of threshold adaption and its effect on the population rate. Since equation \eqref{NO_dyn_approx2} contains a single variable $r_{exc.pop.}(t)$ for the mean excitatory activity, we wanted to find an appropriate description containing the mean threshold $V_{t,pop.}(t) = <V_t(t)>$. As stated earlier, Figure \ref{rate_th_close} suggests an immediate and approximately linear relation between excitatory population rate and the mean of thresholds within the excitatory population. Figure \ref{thresh_r_linfit} shows a linear fit (being of the form $\alpha x + \beta$) of these two quantities plotted against each other. Apart from rarely appearing high values of $V_{t,pop.}$, we found a good fit to the linear model. However, when we color coded the slope of $V_{t,pop.}(t)$, one could see a separation of the rising phase and falling phase within the cloud of data points. This indicates a small (due to the "flatness" of the shape) but noticeable delay between threshold and firing rate, the latter following the former. We implemented this in a similar way as in Equation \eqref{2d-exc_inh_rate_1}. The population rate relaxates towards a value given by the fitted linear dependence on $V_{t,pop.}$ linear fit: $\dot{r_{exc.pop.}}(t) = (-r_{exc.pop.} + r_{IP} + \alpha(V_{t,pop.}(t)-V_{t,0}))/\tau_r$, where $r_{IP}$ is given in table \ref{Params_IP} as the mean excitatory target firing rate and $V_{t,0} \equiv (r_{IP}-\beta)/\alpha$ is the mean threshold corresponding to the target rate. We expect $\tau_r$ to be in the same order of magnitude as the membrane time constant $\tau_m$. However, $r_{exc.pop.}$ represents the whole excitatory population. Taking conduction delay and interactions with the inhibitory population into account, $\tau_r$ thus should be slightly larger.

\begin{figure}
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/linfit_th_rate.png}
\caption{Linear fit of excitatory population activity and mean excitatory threshold. Slope $\alpha$ of the fit: $\mathrm{-0.772\, Hz/mV}$. Offset $\beta$: $\mathrm{-41.01\, Hz}$. Mean squared error: $\mathrm{R^2 = 0.69}$. Colors encode the time derivative of $<V_t(t)>$, which indicates an ellipsoid like trajectory in phase space, with the firing rate following the threshold with a short delay.}
\label{thresh_r_linfit}
\end{figure}

Combining these results in a set of equations, we find
\dot{NO}(t) &= - \lambda NO + \frac{ N_{exc.} {Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3 L^2}r_{exc.pop.} \\
&+ \frac{{Ca^{2+}_{spike}}^3 {\tau_{Ca^{2+}}} ln(2)}{3L^2}\sqrt{\frac{N_{exc.}}{2 \tau_{nNOS}}}\sqrt{r_{exc.pop.}}\xi (t)
\end{split}\label{2d-sys-homeostasis1}\\
\dot{V_{t,pop.}}(t) &= \frac{NO-NO_0}{NO_0 \tau_{V_t}} \label{2d-sys-homeostasis2} \\
\dot{r_{exc.pop.}}(t) &= \frac{-r_{exc.pop.}+\alpha(V_{t,pop.}(t)-V_{t,0}) + r_{IP}}{\tau_r}\;. \label{2d-sys-homeostasis3}
Through the coordinate transformation $n = NO-NO_0 \; , \theta = V_{t,pop.}(t)-V_{t,0}\; , \nu = r_{exc.pop.} - r_{IP}$ we can simplify the dynamical system:
\dot{n} &= -\lambda n + \frac{N_{exc.} {Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3 L^2}\nu \\
&+ \frac{{Ca^{2+}_{spike}}^3 {\tau_{Ca^{2+}}} ln(2)}{3L^2}\sqrt{\frac{N_{exc.}}{2 \tau_{nNOS}}}\sqrt{\nu + r_{IP}}\xi (t)
\end{split}\label{2d-sys-homeostasis-coord_transf1}\\
\dot{\theta} &= \frac{1}{NO_0 \tau_{V_t}} n \label{2d-sys-homeostasis-coord_transf2} \\
\dot{\nu} &= \frac{-\nu + \alpha \theta}{\tau_r} \label{2d-sys-homeostasis-coord_transf3}
where we have used the fact that $NO_0$ can be solved and expressed in terms of $r_{IP}$ by setting $\dot{NO}(t) = 0$. The amplitude of noise in equation \ref{2d-sys-homeostasis-coord_transf1} depends on $\nu$. Keeping this dependence in the equation turned out be an insuperable obstacle for our goal of finding an analytic description of the power spectrum of homeostatic oscillations. In consequence, we decided to take one further step of simplification by replacing $\sqrt{\nu + r_{IP}}$ by a constant factor $\sqrt{r_{IP}}$. In the following equation we use the substitutes $A\equiv \frac{N_{exc.} {Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3 L^2}$, $B\equiv \frac{{Ca^{2+}_{spike}}^3 {\tau_{Ca^{2+}}} ln(2)}{3L^2}\sqrt{\frac{N_{exc.}}{2 \tau_{nNOS}}}\sqrt{r_{IP}}$ and $C\equiv \frac{1}{NO_0 \tau_{V_t}}$, resulting in
\dot{n} &= -\lambda n + A \nu + B \xi (t) \label{2d-sys-homeostasis-coord_transf_substitute1} \\
\dot{\theta} &= Cn \label{2d-sys-homeostasis-coord_transf_substitute2} \\
\dot{\nu} &= \frac{-\nu + \alpha \theta}{\tau_r} \label{2d-sys-homeostasis-coord_transf_substitute3}
As stated in the beginning, we were interested in finding an analytic expression for the power spectrum of the mean excitatory threshold. Therefore we took the fourier transform of \eqref{2d-sys-homeostasis-coord_transf_substitute1} and \eqref{2d-sys-homeostasis-coord_transf_substitute2}, yielding
i\omega f_n &= - \lambda f_n + A f_\nu + B f_\xi \label{2d-sys-homeostasis-coord_transf_substitute_ft1}\\
i\omega f_\theta &= C f_n \label{2d-sys-homeostasis-coord_transf_substitute_ft2} \\
i\omega f_\nu &= \frac{-f_\nu + \alpha f_\theta}{\tau_r} \label{2d-sys-homeostasis-coord_transf_substitute_ft3}
where $f_{(\cdot)}$ denotes the fourier transform. We solved for $f_\theta$ which immediately gave us the power spectrum $P_\theta (\omega) \equiv |f_\theta (\omega)|^2$ of $\theta$:
P_\theta (\omega) = \frac{B^2 P_\xi (\omega)}{\omega^2 \left( \frac{\lambda}{C} + \frac{A\alpha \tau_r}{1+\omega^2 \tau_r^2}\right)^2  + \left(         \frac{A\alpha}{1+\omega^2\tau_r^2} + \frac{\omega^2}{C}\right)^2}
It was already mentioned that the statistics of $\xi(t)$ are approximately gaussian. However, this does not necessarily imply that its power spectrum is white. To find the actual power spectrum, we recall that Equation \eqref{nNOS_single_sol_approx2} can be written as a convolution:
nNOS(t) &= \left(f \ast g\right)(t) \label{Noise_Pow_Spec_Conv1_1} \\
f(t') &= \sum_i \delta(t' - t^i_{spike}) \label{Noise_Pow_Spec_Conv1_2} \\
g(t') &= \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3\tau_{nNOS}} \theta(t') exp(-t'/\tau_{nNOS}) \label{Noise_Pow_Spec_Conv1_3}
Thus $\xi(t)$ is generated in the same way, namely a convolution of an exponential kernel with the same relaxation constant $\tau_{nNOS}$ and a poisson process with a rate of $N_{exc.}\cdot r_{IP}$, since it incorporates spikes from \emph{all} excitatory neurons:
\xi (t) &= \left(f_\xi \ast g_\xi \right)(t) \label{Noise_Pow_Spec_Conv2_1} \\
f_\xi (t') &= \sum_i \sum_{j=1}^{N_{exc.}} \delta(t' - t^{i,j}_{spike}) \label{Noise_Pow_Spec_Conv2_2} \\
g_\xi (t') &= \sqrt{\frac{2}{r_{IP}N_{exc.}\tau_{nNOS}}} \theta(t') exp(-t'/\tau_{nNOS}) \label{Noise_Pow_Spec_Conv2_3}
where $\sqrt{\frac{2}{r_{IP}N_{exc.}\tau_{nNOS}}}$ accounts for the fact that $\xi(t)$ had a variance of one by definition. We can express the power spectrum as
P_{\xi} = P_{f_\xi} \cdot P_{g_\xi} \;. \label{Noise_Pow_Spec_Conv2}
Similar to the earlier assumption that the width of the noise term is stationary, we assume that $f_\xi$ is a stationary Poisson Process with an average rate $r_{IP}N_{exc.}$. It thus has a power spectrum of (see e.g. \cite[p. 162]{Spiking_Gerstner})
P_{f_\xi} (\omega) = r_{IP}N_{exc.} + 2\pi (r_{IP}N_{exc.})^2 \delta(\omega) \; . \label{Pow_Spec_Poisson}
for $P_{g_\xi}$, one gets
P_g(\omega) &= \frac{2}{r_{IP}N_{exc.}\tau_{nNOS}}  \left| \int_{-\infty}^{\infty} \theta(t) exp(-t/\tau_{nNOS}) exp(-i\omega t) dt \right|^2\\
 &= \frac{2}{r_{IP}N_{exc.}\tau_{nNOS}} \frac{1}{\frac{1}{\tau_{nNOS}^2} + \omega^2}
 \end{split} \;. \label{Pow_Spec_Exponential_Kernel}
Since $\xi(t)$ only accounts for the fluctuations around the mean, we can remove the delta function in $P_{f_\xi}$, and get
P_{\xi} = \frac{2}{\tau_{nNOS}} \frac{1}{\frac{1}{\tau_{nNOS}^2} + \omega^2}
Substituting all the intermediate results, we get
\small
P_\theta(\omega) = \frac{1}{N_{exc.}r_{IP}\tau_{nNOS}^2 \left( \frac{1}{\tau_{nNOS}^2} +\omega^2 \right)\left[ \omega^2 \left( \tau_{V_t} + \frac{\alpha \tau_r}{r_{IP}(1+\omega^2\tau_r^2)}\right)^2 + \left( \omega^2 \frac{\tau_{V_t}}{\lambda} + \frac{\alpha}{r_{IP}(1+\omega^2\tau_r^2)} \right)^2 \right] } \; .\label{Pow_Spec_Theta_Final}
\normalsize

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{../../plots/power_spec/power_spectrum_analytic.png}
\end{center}
\caption{Power Spectrum of $\mathrm{\langle V_t \rangle}$ (blue) and analytic predictions based on Equation \eqref{Pow_Spec_Theta_Final} for different time constants of spike rate adaption. Red curve is a fit by tweaking $\tau_m$,$\alpha$ and the overall amplitude of \eqref{Pow_Spec_Theta_Final}.}
\label{Pow_Spec_Theta_vs_Analytic}
\end{figure}
